% R bootcamp, Module 4: Calculations
% August 2014, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
require(foreign)
require(fields)
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
vote <- read.dta('../data/2004_labeled_processed_race.dta')
earnings <- read.dta('../data/heights.dta')
```

# Vectorized calculations

As we've seen, R has many functions that allow you to operate on each element of a vector all at once.

```{r}
vals <- rnorm(1000)
chi2vals <- vals^2
chi2_df1000 <- sum(chi2vals)
# imagine if the code above were a loop, or three separate loops
```

Advantages:

* much faster than looping
* easier to code
* easier to read and understand the code

Sometimes there are surprises in terms of what is fast, as well as tricks for vectorizing things in unexpected ways:
```{r}
vals <- rnorm(1e6)
system.time(trunc <- ifelse(vals > 0, vals, 0))
system.time(vals <- vals * (vals > 0))
```

```{r}
tmp <- as.character(vote$age60)
vote$ageMin <- substring(tmp, 1, 2)
vote$ageMin[1:5]
```

**Question**: What am I doing with `vals * (vals > 0)` ? What happens behind the scenes in R?

If you use a trick like this, having a comment in your code is a good idea.

# Linear algebra 

R can do essentially any linear algebra you need. It uses system-level packages called BLAS (basic linear algebra subroutines) and LAPACK (linear algebra package). Note that these calculations will be essentially as fast as if you wrote C code because R just calls C and Fortran routines to do the calculations.

The BLAS that comes with R is fairly slow. It's possible to use a faster BLAS, as well as one that uses multiple cores automatically. This can in some cases give you an order of magnitude speedup if your work involves a lot of matrix manipulations/linear algebra. More details in Module 10.


# Vectorized vector/matrix calculations

Recall that `+`, `-`,`*`, `/` do vectorized calculations:

```{r}
A <- matrix(1:9, 3)
B <- matrix(seq(4,36, by = 4), 3)

A + B
A + B[ , 1]
A * B
A * B[ , 1]
```

Matrix/vector multiplication

```{r}
A %*% B[ , 1]
A %*% B

identical(t(A)%*%A, crossprod(A))
```

Some decompositions

```{r cache=TRUE}
## next 3 lines generate a positive definite matrix
require(fields)
times <- seq(0, 1, length = 100)
R <- exp(-rdist(times) / 0.2) # a correlation matrix
######################################################
e <- eigen(R)
range(e$values)
e$vectors[ , 1]

sv <- svd(R)
U <- chol(R)

devs <- rnorm(100)
Rinvb <- solve(R, devs)  # R^{-1} b
Rinv <- solve(R) # R^{-1} -- try to avoid this
```


# Pre-allocation

This is slow.
```{r cache=TRUE}
vals <- 0
n <- 50000
system.time({
for(i in 1:n)
      vals <- c(vals, i)
})
```

The same holds for using `rbind()`, `cbind()`, or adding to a list, one element at a time.

This is slow and unclear (in the sense that it appears the code has a bug, but it works):
```{r cache=TRUE}
vals <- 0
n <- 50000
system.time({
for(i in 1:n)
      vals[i] <- i
})
```

**Question**: Thoughts on why these are so slow? Think about what R might be doing behind the scenes

# The answer is to pre-allocate memory

This is not so slow. (Please ignore the for-loop hypocrisy and the fact that I could do this as `vals <- 1:n`.)

```{r}
n <- 50000
system.time({
vals <- rep(0, n)
# alternatively: vals <- as.numeric(NA); length(vals) <- n
for(i in 1:n)
      vals[i] <- i
})
```

Here's how to pre-allocate an empty list: 
```{r}
vals <- list(); length(vals) <- n
head(vals)
```

# apply

Some functions aren't vectorized, or you may want to use a function on every row or column of a matrix/data frame, every element of a list, etc.

For this we use the `apply()` family of functions.

```{r}
mat <- matrix(rnorm(100*1000), nr = 100)
row_min <- apply(mat, MARGIN = 1, FUN = min)
col_max <- apply(mat, MARGIN = 2, FUN = max)
```

There are actually some even faster specialized functions:
```{r}
row_mean <- rowMeans(mat)
col_sum <- colSums(mat)
```

# `lapply()` and `sapply()`

```{r}
myList <- list(rnorm(3), rnorm(3), rnorm(5))
lapply(myList, min)
sapply(myList, min)
```

Note that we don't generally want to use `apply()` on a data frame. 

You can use `lapply()` and `sapply()` on regular vectors, such as vectors of indices, which can come in handy, though this is a silly example:
```{r}
sapply(1:10, function(x) x^2)
```

Here's a cool trick to pull off a particular element of a list of lists:

```{r}
params <- list(a = list(mn = 7, sd = 3), b = list(mn = 6,sd = 1), 
  c = list(mn = 2, sd = 1))
sapply(params, "[[", 1)
```

**Challenge**: Think about why this works. 

Hint:
```{r}
test <- list(5, 7, 3)
test[[2]]
# `[[`(test, 2)  # need it commented or R Markdown processing messes it up...

# `+`(3, 7)
```

# And more `apply()`s

There are a bunch of `apply()` variants, as well as parallelized versions of them:

* `tapply()`, `vapply()`, `mapply()`, `rapply()`, `eapply()`
* for parallelized versions see Module 10 or `?clusterApply`)

# Tabulation 

- Sometimes we need to do some basic checking for the number of observations or types of observations in our dataset
- To do this quickly and easily, `table()` is our friend
- Let's look at our observations by year and grade

```{r table}
unique(vote$pres04)
tbl <- table(vote$race, vote$pres04)
tbl
round(prop.table(tbl, margin = 1), 3)
table(vote$race, vote$pres04, vote$sex)
with(vote[vote$sex == 'female', ], table(pres04, race))
```

**Challenge**: Can you figure out what `with()` does just by example? 

# Stratified analyses I
Often we want to do individual analyses within subsets or clusters of our data.

As a first step, we might want to just split our dataset by a stratifying variable.

```{r}
subsets <- split(earnings, earnings$race)
length(subsets)
subsets[['9']]
```

The `%in%` operator can also be helpful.

```{r}
sub <- earnings[earnings$race %in% c(1, 2, 4), ]
```

# Stratified analyses II

Often we want to do individual analyses within subsets or clusters of our data. R has a variety of tools for this; for now we'll look at `aggregate()` and `by()`. These are wrappers of `tapply()`. 

```{r aggregate1}
aggregate(earnings, by = list(educ = earnings$ed), FUN = median, na.rm = TRUE)
aggregate(earn ~ ed, data = earnings, FUN = median)
aggregate(earn ~ ed + hisp, data = earnings, FUN = median)
agg <- aggregate(earn ~ ed + hisp, data = earnings, FUN = median)
xtabs(earn ~ ., data = agg)
```

Notice the 'long' vs. 'wide' formats. You'll see more about that sort of thing in Module 5.

# Discretization

You may need to discretize a continuous variable [or a discrete variable with many levels], e.g., by education level:
```{r fig.width=9}
earnings$edLevel <- cut(earnings$ed, breaks = c(0, 11, 12, 15, 16, 18, 99))
levels(earnings$edLevel) <- c('no HS diploma', 'HS grad', 'some college', 
   'college grad', 'grad study', 'other')
boxplot(earn~edLevel, data = earnings)
```


# Stratified analyses III

`aggregate()` works fine when the output is univariate, but what about more complicated analyses than computing the median, such as fitting a set of regressions?

```{r}
out <- by(earnings, earnings$edLevel, 
function(x) {
  if(sum(!is.na(x$earn))) 
    lm(earn ~ height, data = x) 
  else NA
})
length(out)
summary(out[[5]])
```

**Question**: What's the business with the `if` statement? Why is this good practice?

# Sorting

`sort()` applied to a vector does what you expect.

Sorting a matrix or dataframe based on one or more columns is a somewhat manual process, but once you get the hang of it, it's not bad.

```{r}
ord <- order(earnings$earn, earnings$height, decreasing = TRUE)
# ord <- with(earnings, order(earn, height, decreasing = TRUE))
ord[1:5]
(earnings$earn[ord])[c(1:5, 2025:2029)] # parentheses for clarity
earnings_ordered <- earnings[ord, ]
```

You could of course write your own *sort* function that uses `order()`. More in Module 6.

# Merging Data

We often need to combine data across multiple data frames, merging on common fields (i.e., *keys*). In database terminology, this is a *join* operation.

Here's an example using the voting data combined with a built-in R dataset on state information. Warning: the state dataset is *very* old; this is just a toy example. 

In this case (as often true) we need to do some machinations to get everything prepared for the merge. The *key* we use is the state name.

```{r} 
# PRE-PROCESSING

# a bit of querying indicates the state numbers are in alphabetical order, 
# so we can do this to get a 'name' column in the 'vote' dataframe:
numToName <- data.frame(stateNum = 1:50, 
   stateName = row.names(state.x77)) 
voteWithStateNames <- merge(vote, numToName, by.x = 'state', 
   by.y = 'stateNum', all.x = TRUE, all.y = FALSE)
stateInfo <- data.frame(state.x77)
# need the names as column, not as the row names attribute:
stateInfo$name <- row.names(stateInfo) 

# ACTUAL DEMONSTRATION
fullVote <- merge(voteWithStateNames, stateInfo[ , 
   c('name', 'Population', 'Income', 'Illiteracy', 'Life.Exp')], 
   by.x  = 'stateName', by.y = 'name', all.x = TRUE, all.y = FALSE)
dim(vote)
dim(fullVote)
fullVote[1:2, ]
```

What's the deal with the `all.x` and `all.y` ?  We can tell R whether we want to keep all of the `x` observations, all the `y` observations, or neither, or both, when there may be rows in either of the datasets that don't match the other dataset.

# Breakout

## Problem 1

Suppose we have two categorical variables and we conduct a hypothesis test of independence. The chi-square statistic is: 

$$
\chi^2 = \sum_{i=1}^{n}\sum_{j=1}^{m} \frac{(y_{ij} - e_{ij})^2}{e_{ij}}, 
$$ 

where $e_{ij} = \frac{y_{i\cdot} y_{\cdot j}}{y_{\cdot \cdot}}$, with $y_{i\cdot}$ the sum of the values in the i'th row, $y_{\cdot j}$ the sum of values in the j'th column, and $y_{\cdot\cdot}$ the sum of all the values. Suppose I give you a matrix in R with the $y_{ij}$ values. 

You can generate a test matrix as: `y <- matrix(sample(1:10, 12, replace = TRUE), nrow = 3, ncol = 4)`.

Compute the statistic without *any* loops as follows:

1. First, assume you have the *e* matrix. How do you compute the statistic without loops as a function of `y` and `e`?
2. How can you construct the *e* matrix? Hint: the numerator of *e* is just an *outer product* for which the `outer()` function can be used.


## Problem 2

For each combination of sex and education level in the *earnings* dataset (*data/heights.dta*), find the *second* largest value of earnings amongst the people in the group without any looping.
