% R bootcamp, Module 11: Breakout solutions
% August 2013, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
require(foreign)
require(fields)
vote <- read.dta('../../data/2004_labeled_processed_race.dta')
```

# Problem 2

Fit logistic regression models of preference for Bush/Kerry on income, stratified by state. Use `foreach` or a parallel version of one of the *apply* variants.  Collect the resulting coefficients (and standard errors, if you're feeling ambitious) in a clean format such as a matrix or data frame.  Check to see if multiple cores are being used in the execution.


```{r}
require(parallel)
nCores <- 4
p <- sum(table(vote$income) != 0)
vote <- vote[vote$pres04 == 1 | vote$pres04 == 2, ]
vote$pres04[vote$pres04 == 2] <- 0 # 0=bush; 1=kerry


spl <- split(vote, vote$state)
system.time({out <- lapply(spl, 
  function(x) {
    mod <- glm(pres04 ~ income, data = x)
    summary(mod)$coefficients[ , 1:2] # leave off z-value/p-value
  }
)})

system.time({out <- mclapply(spl, 
  function(x) {
    mod <- glm(pres04 ~ income, data = x)
    summary(mod)$coefficients[ , 1:2] # leave off z-value/p-value
  }
)})
coefs <- array(unlist(out), c(p, 2, length(spl)))
identical(c(coefs[ , , 3]), c(out[[3]])) # check ordering...

```

Note that there was no point in parallelizing this particular computation given the size of the problem!

