% R bootcamp, Module 5: Breakout solutions
% August 2014, UC Berkeley
% Chris Krogslund

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
```

# Breakout and overnight homework

Consider the tipping data. Fit a logistic regression to the data for SUNDAY, modeling the probability of leaving a good tip (>=15%) as a function of the total bill and the size of the party, potentially including additional covariates such as sex, smoker status, and time. What do you find in terms of how the total bill associates with the probability of leaving a good tip?

How do you predict the probability of leaving a good tip for a given set of covariate values? Consider the `predict.glm()` function and what its help page says. Or write code that converts from the model coefficients to the probability scale. Compare the predicted probability of leaving a good tip for a party of two with a party of four.

# Breakout Solutions

```{r}
library(reshape2)

# Load data
data(tips)

# Create binary indicator for good tip (i.e. >=15%)
tips$good.tip<-ifelse(test = tips$tip/tips$total_bill>=0.15, yes = 1, no = 0)

# Subset the data to include just observations from Sundays
tips.sun<-tips[tips$day=="Sun",]

# Regress good.tip on total_bill and size; get summary output; isolate total_bill coefficient
reg<-glm(formula = good.tip~total_bill+size, family = binomial(logit), data = tips.sun)
sum.reg<-summary(reg)
sum.reg$coefficients

# Plot predicted values versus total_bill
plot(x = tips.sun$total_bill, y = exp(predict(object = reg)), xlab = "Total Bill", ylab = "Pr(Good Tip)")

# Re-estimate model with interactions
reg<-glm(formula = good.tip~total_bill*size, family = binomial(logit), data = tips.sun)
sum.reg<-summary(reg)
sum.reg$coefficients

# Plot predicted values versus total_bill
plot(x = tips.sun$total_bill, y = exp(predict(object = reg)), xlab = "Total Bill", ylab = "Pr(Good Tip)")

# Re-estimate model with quadratic term on total_bill
reg<-glm(formula = good.tip~I(total_bill^2)+total_bill+size, family = binomial(logit), data = tips.sun)
sum.reg<-summary(reg)
sum.reg$coefficients

# Plot predicted values versus total_bill
plot(x = tips.sun$total_bill, y = exp(predict(object = reg)), xlab = "Total Bill", ylab = "Pr(Good Tip)")

# Re-estimate model with additional covariates
reg<-glm(formula = good.tip~total_bill+size+sex+smoker, family = binomial(logit), data = tips.sun)
sum.reg<-summary(reg)
sum.reg$coefficients

# Plot predicted values versus total_bill
plot(x = tips.sun$total_bill, y = exp(predict(object = reg)), xlab = "Total Bill", ylab = "Pr(Good Tip)")

# Re-estimate model with interactions and additional covariates
reg<-glm(formula = good.tip~total_bill*size+sex+smoker, family = binomial(logit), data = tips.sun)
sum.reg<-summary(reg)
sum.reg$coefficients

# Plot predicted values versus total_bill
plot(x = tips.sun$total_bill, y = exp(predict(object = reg)), xlab = "Total Bill", ylab = "Pr(Good Tip)")

# Re-estimate model with quadratic term on total_bill and additional covariates
reg<-glm(formula = good.tip~I(total_bill^2)+total_bill+size+sex+smoker, family = binomial(logit), data = tips.sun)
sum.reg<-summary(reg)
sum.reg$coefficients

# Plot predicted values versus total_bill
plot(x = tips.sun$total_bill, y = exp(predict(object = reg)), xlab = "Total Bill", ylab = "Pr(Good Tip)")

# Compare predicted probability of leaving a good tip for group sizes of two and four
plot(x = tips.sun$size, y = exp(predict(object = reg)), xlab = "Size of Party", ylab = "Pr(Good Tip)")
mean(exp(predict(object = reg))[tips.sun$size==2])
mean(exp(predict(object = reg))[tips.sun$size==4])
```

# Ambitious Homework

Using the tools for stratified analyses we have seen today, fit separate models of tipping behavior across different days. How do the effects of the total bill amount and size of party vary by day?

For our purposes here, don't worry much about the uncertainty in the estimates you get in the logistic regression modeling, but of course in a real analysis you would need to do this. 

# Ambitious Solutions
```{r}
# Compare effects of tota_bill and party size across days
out<-lapply(X = unique(tips$day), FUN = function(day){
  reg<-glm(formula = good.tip~total_bill+size+sex+smoker, family = binomial(logit), data = tips[tips$day==day,])
  data.frame(summary(reg)$coefficients, day=day, var=rownames(summary(reg)$coefficients))
})
out<-do.call("rbind", out)
out[out$var=="total_bill",]
out[out$var=="size",]
```










