% R bootcamp, Module 1: Breakout solutions
% August 2014, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
require(foreign)
require(fields)
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
vote <- read.dta('../data/2004_labeled_processed_race.dta')
earnings <- read.dta('../data/heights.dta')
```

# Problem 1

Create a new R object that contains the heights in meters ($1 m \approx 39.4 in$) using the *height1* and *height2*  variables in the *earnings* dataset. Make a histogram of the new variable with breakpoints between the bins every 10 cm.


```{r}
inchPerMeter <- 39.4
inchPerFoot <- 12
hgt_m <- (earnings$height1 * inchPerFoot + earnings$height2) / inchPerMeter
hist(hgt_m, breaks = seq(1.4, 2.1, by = .1))
```

# Problem 2

For a random subset of 100 individuals from the earnings dataset, create a vector with their earnings.


```{r}
smp <- sample(1:nrow(earnings), 100)
sub <- earnings$earn[smp]
```



