% R bootcamp, Module 4: Breakout solutions
% August 2014, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
require(foreign)
require(fields)
vote <- read.dta('../../data/2004_labeled_processed_race.dta')
earnings <- read.dta('../../data/heights.dta')
```

# Problem 1

**To see this problem with the equations below rendered correctly, see module3_calc.html (the non-slide version).**

Suppose we have two categorical variables and we conduct a hypothesis test of independence. The chi-square statistic is: 

\[
\chi^2 = \sum_{i=1}^{n}\sum_{j=1}^{m} \frac{(y_{ij} - e_{ij})^2}{e_{ij}}, 
\] 

where $e_{ij} = \frac{y_{i\cdot} y_{\cdot j}}{y_{\cdot \cdot}}$, with $y_{i\cdot}$ the sum of the values in the i'th row, $y_{\cdot j}$ the sum of values in the j'th column, and $y_{\cdot\cdot}$ the sum of all the values. Suppose I give you a matrix in R with the $y_{ij}$ values. Compute the statistics without *any* loops.

You can generate a test matrix as: `y <- matrix(sample(1:10, 12, replace = TRUE), nrow = 3, ncol = 4)`.

1. Assume you have the *e* matrix. How do you compute the statistic without loops?
2. How can you construct the *e* matrix? Hint: the numerator of *e* is just an *outer product* for which the `outer()` function can be used.


```{r}
        y <- matrix(sample(1:10, 12, replace = TRUE), nrow = 3, ncol = 4)     
     yidot <- rowSums(y)
     ydotj <- colSums(y)
     e <- outer(yidot, ydotj) / sum(y)
     sum( (y - e)^2 / e)
```


# Problem 2

For each combination of sex and education level, find the *second* largest value of earnings amongst the people in the group without any looping.


```{r}
results <- by(earnings, list(earnings$sex, earnings$ed), 
  function(x) {
  tmp <- sort(x$earn, decreasing = TRUE)
  tmp[2]
  }
)
```

Hmm, that format is awkward. 

Here's a bit of code to get it in a less awkward format.

```{r}
dim(results)
results[1, ]
matrix(c(results), length(unique(earnings$sex)), length(unique(earnings$ed)))
```

Are there other ways to go about this? 

This is cleaner, but misrepresents the missing values. 

```{r}
agg <- aggregate(earn ~ sex + ed, data = earnings, FUN = function(x) {
   sort(x, decreasing = TRUE)[2]
})
xtabs(earn ~ ., data = agg)

```


