% R bootcamp, Module 8: Breakout solutions
% August 2014, UC Berkeley
% Chris Krogslund (ckrogslund@berkeley.edu)

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(ggplot2)
data<-read.csv("../../data/cpds.csv")
```

# Breakout Questions:

1) Not all variable types are suitable for representation by every ggplot aesthetic.  What kinds of variables can the aesthetics color, size, and shape meaningfully represent? 

2) Using ggplot2, create a trellis plot where, for a given country, each panel uses a) HOLLOW CIRCLES to plot real GDP growth over time, and b) a red LOESS smoother without standard errors to plot the trend in unemployment over time.  BONUS: limit the years shown to the period from 2000 to 2010, and turn off the grey background.

# Breakout Answers:

1) 

Color: discrete, continuous, ordered, unordered; 
Size: discrete, continuous, ordered, unordered; 
Shape: discrete, unordered

2) 

```{r}
library(ggplot2)

# Create plot
plot<-ggplot(data=data)+
  geom_point(aes(x = year, y = realgdpgr),shape=1)+
  geom_smooth(aes(x = year, y = unemp), se=F, method="loess", color="red")+
  facet_wrap(facets = ~country)
plot

# Limit x-axis range and turn off grey background
plot+xlim(2000,2010)+theme_bw()
```


