% R bootcamp, Module 8: Breakout solutions
% August 2013, UC Berkeley
% Chris Krogslund (ckrogslund@berkeley.edu)

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(plyr)
library(reshape2)
library(ggplot2)
library(lattice)
library(gridExtra)
mydata<-read.csv("data/cpds.csv")
```

Breakout Questions:

1) Not all variable types are suitable for representation by every ggplot aesthetic.  What kinds of variables can the aesthetics color, size, and shape meaningfully represent? 

2) ggplot graphics are often layered, with some data represented by a set of aesthetics being overlayed upon a different combination of data and aesthetics.  Can you fit simple linear trend lines to every facet of the grided voter turnout graph?  Hint: use a simple linear regression and consult ?geom_smooth

Breakout Answers:

1) 

Color: discrete, continuous, ordered, unordered; 
Size: discrete, continuous, ordered, unordered; 
Shape: discrete, unordered

2) 

```{r}
ggplot(data=mydata, aes(x=year, y=vturn))+geom_point()+facet_wrap(~country)+
  geom_smooth(method="lm", color="red", size=1)
```


