% R bootcamp, Module 3: Breakout solutions
% August 2014, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
require(foreign)
require(fields)
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
vote <- read.dta('../data/2004_labeled_processed_race.dta')
earnings <- read.dta('../data/heights.dta')
answers <- c(TRUE, TRUE, FALSE, FALSE)
update <- c(TRUE, FALSE, TRUE, FALSE)
```

# Problem 1

Using the presidential preference dataset, create a new column based on *age9* that gives, as a numeric value, the midpoint of the age range assigned to each person. Try to do this with a combination of subsetting and string operations (i.e., can you convert the character numbers to actual numbers). To simplify things, feel free to get rid of the rows for ages "75 and over".
    

```{r}
vote$age9 <- as.character(vote$age9)
vote$age9[vote$age9 == "75 or over"] <- "75-99"
lower <- as.numeric(substring(vote$age9, 1, 2))
upper <- as.numeric(substring(vote$age9, 4, 5))
vote$approxAge <- (lower + upper) / 2
hist(vote$approxAge, breaks = 40)
```

# Problem 2

Go back to slide 6 on logical vectors and figure out what is going on in the last few lines of code.


```{r}
answers & update
answers * update
as.numeric(answers)*as.numeric(update)

answers | update
answers + update
as.logical(answers + update)
```

# Problem 3
Go back to slide 9 and explain why `vote[c(3,5)]` and `vote[ , c(3,5)]` produce the same result.

```{r}
class(vote)
is.list(vote)
identical(vote[1], vote[ , 1, drop = FALSE])
class(vote[c(3,5)])
```
