% R bootcamp, Module 5: Programming
% August 2013, UC Berkeley
% Jacob Lynn
========================================================

# Solutions to breakout problems: problem #1

1) Write a function that returns the mean of N normally distributed random numbers.

```{r}
meanOfNormals <- function(N) {
  mean(rnorm(N))
}

meanOfNormals(1)
meanOfNormals(10)
meanOfNormals(100)
meanOfNormals(1000)
meanOfNormals(10000)
```

# Solutions to breakout problems: problem #2

2) Extend exercise #1: allow the user to pass parameters for the normal distribution, and return both mean and median.

```{r}
statsOfNormals <- function(N, mean=0, sd=1) {
  myVars <- rnorm(N, mean, sd)
  out <- list(mean(myVars), median(myVars))
  names(out) <- c("mean","median")
  out
}

statsOfNormals(1, 5, 2)
statsOfNormals(10, 5, 1)
statsOfNormals(100, 3)
statsOfNormals(1000, 3)
statsOfNormals(10000, 3)
```

# Solutions to breakout problems: problem #3

3) Write a function that repeatedly generates random normal variables until it generates a random number more than N standard deviations from the mean. Return the number of samples performed up to that point.

```{r}
nUntilFluctuation <- function(N) {
  i <- 0
  var <- 0
  while (abs(var) < N) {
    var = rnorm(1)
    i = i+1
  }
  i
}

nUntilFluctuation(1.0)
nUntilFluctuation(2.0)
nUntilFluctuation(3.0)
nUntilFluctuation(4.0)
```
